---
title: "week 10"
format:
  html:
    embed-resources: true
editor: visual
---

## Homework

Calculate odds ratio of 4 variables.

```{r}
#| message: false
#| warning: false
#| echo: false
#| cashe: false

library(gssr)
library(tidyverse)

```

## Variale 1

What's the odds ratio of having a college degree being a woman?

```{r}
gss2018 <- gss_get_yr(2018)
gendereduc <- gss2018 |>
  select(sex, educ) |>
  mutate(gender = if_else(sex == 1 , 1, 0),
         degree = if_else(educ >= 16, 1, 0)) |>
  drop_na()
```

Outcome: having college degree(Yes=1 / No=0)

Predictor:gender (Men=1 / Women=0)

```{r}
table(gendereduc$gender, gendereduc$degree)
```

Probability of being a woman $P(gender = 0)$ is `r 867/(867+715)`

Probability of being a man $P(gender = 1)$ is `r 715/(867+715)`

Odds of being a woman are: 

$$
\frac{P(gender=0)}{1- P(gender=0)} = \frac{0.5480}{1-0.5480} = 1.21
$$ 

Probability of having a college degree in the condition of being a woman $P(degree = 1 | gender = 0)$ is `r 427/(427+867)`

Probability of having a college degree in the condition of being a man $P(degree = 1 | gender = 1)$ is `r 336/(715+336)`

Odds ratio of two odds: 

$$
OR = \frac{\frac{P(degree = 1| gender = 0)}{1-P(degree =1|gender = 0)}} {\frac{P(degree = 1| gender = 1)}{1-P(degree =1|gender=1)}}
$$ 

Odds ratio = (0.33/(1-0.33))/(0.32/(1-0.32)) = `r (0.33/(1-0.33))/(0.32/(1-0.32))`

#### Writing as a model

$$
log(\frac{P(degree = 1)}{1-P(degree =1)}) = \alpha + \beta(gender_i)
$$

```{r}
mod1 <- glm(degree ~ gender,
            data = gendereduc,
            family = binomial(link = "logit"))
summary(mod1)
```

$$
log(\frac{P(degree = 1)}{1-P(degree =1)}) = -.708-.047*gender
$$

```{r}
exp(-.047)
```

OR = .9541 When gender changes from women to men, the odds of having a college degree decrease by 4.6%.

## Variable 2

What is the odds ratio of late first birth age for those with a college degree? We define late first birth as having a first child at age 28 or older.

```{r}
firstkid <- gss2018 |>
  select(agekdbrn, educ) |>
  mutate(latebirth = if_else(agekdbrn >= 28, 1, 0), 
         degree = if_else(educ >= 16, 1, 0)) |>
  drop_na()
```

Outcome: late first birth age (Yes=1 / No=0)

Predictor:having college degree(Yes=1 / No=0)

```{r}
table(firstkid$degree, firstkid$latebirth)
```

Probability of having late first birth in the condition of having a college degree $P(latebirth = 1 | degree = 1)$ is `r 253/(248+253)`

Probability of having late first birth in the condition of not having a college degree $P(latebirth = 1 | degree = 0)$ is `r 205/(959+205)`

OR = (.5050/(1-.5050))/(/.1761/(1-.1761)) = `r (.5050/(1-.5050))/(.1761/(1-.1761))`

#### Writing as a model

$$
log(\frac{P(latebirth = 1)}{1-P(latebirth =1)}) = \alpha + \beta(degree_i)
$$

```{r}
mod2 <- glm(latebirth ~ degree,
            data = firstkid,
            family = binomial(link = "logit"))
summary(mod2)
```

$$
log(\frac{P(latebirth = 1)}{1-P(latebirth =1)}) = -1.5429 + 1.5628*degree
$$

```{r}
exp(1.5628)
```

OR = 4.77, the odds of late first birth having a college degree is 4.77 times greater than the odds of late first birth not having a college degree. 

## Variable 3

What's the odds ratio of having a college degree of people who born in the U.S compared to those not?

```{r}
citi <- gss2018 |>
  select(born, degree) |>
  mutate(BornUS = if_else(born == 1, 1, 0),
         college = if_else(degree >= 3, 1, 0)) |>
    drop_na() 
```

Outcome: college degree (Yes=1 / No=0)

Predictor:whether born in U.S. (Yes=1 / No=0)

```{r}
table(citi$BornUS, citi$college)
```

Probability of having college degree in the condition of born in US $P(college = 1 | BornUS = 1)$ is `r 619/(1426+619)`

Probability of having college degree in the condition of not born in US $P(college = 1 | BornUS = 0)$ is `r 93/(209+93)`

OR = (.3027/(1-.3027))/(.3079/(1-.3079)) = `r (.3027/(1-.3027))/(.3079/(1-.3079))`

#### Writing as a model

$$
log(\frac{P(college = 1)}{1-P(college =1)}) = \alpha + \beta(BornUS_i)
$$

```{r}
mod3 <- glm(college ~ BornUS,
            data = citi,
            family = binomial(link = "logit"))
summary(mod3)
```

$$
log(\frac{P(college = 1)}{1-P(college =1)}) = -.8097 - .0248*BornUS
$$

```{r}
exp(-.0248)
```

OR = .9755. It indicates that the odds ratio of having a college degree decreases by 2.4% when they are born in the U.S compared to those not.

## Variable 4

What's the odds ratio of people having college degree if their father has a college degree compared to those not?

```{r}
fathered <- gss2018 |>
  select(paeduc, degree) |>
  mutate(Pacollege = if_else(paeduc >= 16, 1, 0),
         college = if_else(degree >= 3, 1, 0)) |>
    drop_na() 
```
Outcome: college degree (Yes=1 / No=0)

Predictor:Father having college degree (Yes=1 / No=0)

```{r}
table(fathered$Pacollege, fathered$college)
```

Probability of having college degree in the condition of father having a college degree $P(college = 1 | Pacollege = 1)$ is `r 228/(155+228)`

Probability of having college degree in the condition of father not having a college degree $P(college = 1 | Pacollege = 0)$ is `r 378/(926+378)`
